---
title: "Aggregating some observations in a data frame"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A common task when working on spatio-temporal data frames is aggregating the
observations of some variables, either in space, or in time. This can be done
easilty with the `aggregate_by` function of the `dplyrx` package. Let's load it:

```{r}
library(dplyrx)
```

To understand it better what the function can help do let's have a  look at the
following meteorological data from the `dplyrx` package:

```{r meteo_data}
data(meteo)
meteo
```

This data set contains the values of temperature (°C), absolute humidity
(g / m$^3$), rainfall (mm) and hours of sunshine from 3 meteorological stations
(Can Tho, Da Lat and Da Nang) for the months of January, May and September in
2013, 2014 and 2015. From here, imagine that you want to aggretate all the
observations of the stations Can Tho and Da Lat. An obvious way to do so is
probably to take the mean. This can be done so in two step with the package 
`dplyr` and with the `aggregate_by` function of the `dplyrx` package:
First, we need to `recode` the stations selected with the same value and then
used the `aggregate_by` function on the output. 

```{r}
library(dplyr)
meteo %>% 
  mutate(station = recode(station, `Can Tho` = "group_prov", `Da Lat`= "group_prov")) %>% 
  aggregate_by(station, month, year, .funs = mean)
```

The second argument are listed all the variables that should be used to 
identify unique observations to aggregate in each of the stations Can Tho and 
Da Lat. Here we want aggregate by space but keep the temporal information 
contained in the variables `month` and `year`. 

The last argument, `.funs` specifies the function that should be used to perform
the aggregation. By default it's `sum` but here it makes more sense to do 
`mean`.

We can naturally aggregate the observations and apply the function on 
`temperature` and `humidity` only, the argument `.funs` need to be specify as a
list:

```{r}
meteo %>% 
  mutate(station = recode(station, `Can Tho` = "group_prov", `Da Lat`= "group_prov")) %>% 
  aggregate_by(station, month, year, .funs = list(mean(temperature, humidity)))
```

## A more complex situation

Let's now consider the case where, with the `meteo` data set, instead of
performing a spatial aggregation we want to  perform a temporal aggregation.
Let's imagine for example that we want to aggregate the months of January and
May for all the provinces. What's apparent here is that the proper way of
performing such an aggregation for the variables `temperature` and `humidity`
would probably to use the `mean` function whereas it would certainly make more
sense to use the `sum` function for the variables `rainfall` and `sunshine`. The
pipeline below shows how to do so:

```{r}
meteo %>% 
  mutate(month = recode(month, `January` = "group_month", `May`= "group_month")) %>% 
  aggregate_by(station, month, year, .funs = list(mean(temperature, humidity), sum(rainfall, sunshine)))
```
